---
title: "OBDS_R_week_1_day_2_afternoon"
author: "Laila Latifi"
date: "2025-05-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Sorting Vectors
sort function only for vectors.
To sort data frames, need to use order and index (including , to do by rows).
Ex. ToothGrowth[c(order(ToothGrowth$len) , )]
In tidyverse, built in functions to get around this and sort matrix.


```{r}
data(ToothGrowth)
ToothGrowth[c(order(ToothGrowth$len), )]
```


# Exercise - Sorting Data Frames

Load the airquality data set.
```{r}
data("airquality")
```

Open the help page for this data set.
```{r}
?airquality
```

Examine the data set.
```{r}
str(airquality)
head(airquality)
View(airquality)
```

Display the column names of the airquality data frame.
```{r}
colnames(airquality)
```

Sort the data frame by increasing value in the Ozone column.
```{r}
order(airquality$Ozone)
airquality[order(airquality$Ozone), ]
```
Note, NAs are all at the end.

Sort the data frame by Month in increasing order and Temp in decreasing order.
```{r}
# inc_month <- order(airquality$Month)
# dec_temp <- order(airquality$Temp, decreasing = TRUE)
# airquality[inc_month, ]
?order

order(airquality$Month, airquality$Temp, decreasing = c(FALSE, TRUE))
airquality[order(airquality$Month, airquality$Temp, decreasing = c(FALSE, TRUE)), ]

airquality_month_temp_sorted <- airquality[
  order(airquality$Month, airquality$Temp, 
  decreasing = c(FALSE, TRUE)), ]

View(airquality_month_temp_sorted)
```

Write the latest sorted data frame to a text file format of your choice.
```{r}
write.table(airquality_month_temp_sorted,
  file = "airquality_month_temp_sorted.tsv",
  sep = "\t", quote = FALSE)

read.delim("airquality_month_temp_sorted.tsv")
```



# Exercise - Merging Data Frames

Run the code below to create two data frames.
```{r}
buildings <- data.frame(
  site = c(1, 2, 3),
  name = c("b1", "b2", "b3"))

survey_data <- data.frame(
  survey = c("A", "A", "A", "B", "B", "B"),
  location = c(1, 2, 3, 2, 3, 1),
  efficiency = c(51, 64, 70, 71, 80, 58))
```

What is the shared information in these two data frames?
```{r}
str(buildings)
str(survey_data)

View(buildings)
View(survey_data)
```
The buildings$site and survey_data$location are the same information. 

Use the merge() function to combine the two data frames by the shared 
information into a new data frame called buildings_survey.
```{r}
buildings_survey <- merge(
  x = buildings, 
  y = survey_data, 
  by.x = "site", 
  by.y = "location")

View(buildings_survey)
```



# Exercise - Summarising Groups of Data

Compute the mean of each numeric column each month in the airquality data frame 
using aggregate(). Make sure NA values are removed.
```{r}
str(airquality)

aggregate(x = airquality, 
          by = list(month = airquality$Month), 
          FUN = mean, 
          na.rm = TRUE)

```

Compute the mean of the Solar.R column each month. Make sure the grouping 
column is called Month in the return value. Make sure NA values are removed.
```{r}
aggregate(x = airquality$Solar.R, 
          by = list(Month = airquality$Month), 
          FUN = mean, 
          na.rm = TRUE)

# Can also write with c()
aggregate(x = airquality[, c("Solar.R", "Temp")], 
          by = list(Month = airquality$Month), 
          FUN = mean, 
          na.rm = TRUE)
```



# User Defined Functions

Message is useful for knowing what your function is doing at a given point in
time will it is running, particularly if the function takes a long time to run.
Message doesn't impact the output.

... means function is accepting extra argument, but not telling you what they 
are. Allows function to capture information from you and pass it down to the 
function. 
Ex. na.rm = TRUE in aggregate function above.
na.rm not an argument in aggregate function, but we still used it to remove
NAs and then pass the remaining information through the rest of the function
to get the mean.

As on slide, can give input values without specifying (i.e. writing 1, 2, 3 
instead of apples = 1, pears = 2, oranges = 3).
Can set objects to 0 by default so function will still run even if all 
possible values are not given.



# Exercise - Writing a Function

Write a function to calculate the hypotenuse of a triangle given the length of 
the other two sides.
```{r}
hypotenuse_length <- function(a, b) {
  hypotenuse <- sqrt(a^2 + b^2)
  return(hypotenuse)
}
```

Run the function you have created with different values.
```{r}
hypotenuse_length(2, 2)
hypotenuse_length(3, 4)

hypotenuse_length(a = 3, b = 4)
hypotenuse_length(b = 3, a = 4)

hypotenuse_length(1)
hypotenuse_length(b = 1)
```

The variables a function defines only exist while the funciton is running. 
Objects a funciton creates internally are discarded afterwards. Only the 
output of the function stays. R "scoping".



# Applying Functions

apply(), MARGIN = 1 for row or 2 for column
Ex. Given under "by row" on slide is same as rowmean() function we used earlier.

sapply() great to use interactively to simplify, but not good to use as a 
developer as letting R do things automatically can lead to unforseen mistakes.

vapply() faster since it knows how to allocate memory ahead of time due to you
specifiying size of inputs and outputs. 
numeric(1) means one numeric value.



# Exercise - Apply

Create a vector of integers from 1 to 10.
```{r}
int_vector <- 1:10
int_vector
```

Compute the log2 of each value in the vector using either lapply() or sapply().
```{r}
?log2 

lapply(X = int_vector, FUN = log2)
sapply(X = int_vector, FUN = log2)

# Or
lapply(int_vector, log2)
sapply(int_vector, log2)
```

Compare the outputs of lapply() and sapply() in the previous step.
lapply() gives everything separately as lists, but sapply() combines all the 
outputs into a vector.

Create a list of four elements, each element being a vector of type either 
numeric or logical.
```{r}
ex_list <- list(1:12, c(TRUE, TRUE), 17:20, c(FALSE, FALSE))
```

Compute the sum of each vector in the list using either lapply() or sapply().
```{r}
lapply(X = ex_list, FUN = sum)
sapply(X = ex_list, FUN = sum)
```

Use sapply() on the list that you created in part 2, to repeat each element of each vector three times. i.e., 1, 2, 3 should become 1, 1, 1, 2, 2, 2, 3, 3, 3
```{r}
?rep
sapply(X = ex_list, FUN = rep, each = 3)
```



# Loops

Will use for loops 99% of the time in bioinformatics.



# Exercise - Loops and Conditions

Write a for loop that iterates over the integers 1 to 7 and prints the number 
raised to the power of three.
```{r}
for (i in 1:7) {
  print(i^3)
}
```

Write a for loop that iterates over the names of the columns in the built in 
data set iris and prints each column name together with the number of 
characters in that column name.

Example output: Sepal.Length: 12
Hint: use the functions print(), paste0(), and nchar().
Remember to read the help page of each function to learn more about them.

My attempt
```{r}
?nchar()
?paste0

data("iris")

head(iris)
str(iris)
colnames(iris)

# Creating character vectors from the column names
paste0(colnames(iris))
colname_vector <- paste0(colnames(iris))

# Count the number of characters in each object in vector
nchar((paste0(colnames(iris))), type = "chars")


for (i in colname_vector) {
  print(i)
}
```

Solution
```{r}
data("iris")

for (column_name in colnames(iris)) {
  # print(column_name)
  # print(nchar(column_name))
  print(paste0(column_name, ": ", nchar(column_name)))
}
```

Use the ifelse() function to print the name of colours that are made up of four 
characters in the vector my_colours below.
```{r}
my_colours <- c("red", "orange", "purple", "yellow", "pink", "blue")

ifelse(test = nchar(my_colours) == 4, yes = my_colours, no = "")
```